Is There a Competitive Advantage to Using Multivariate Statistical or Machine Learning Methods Over the Bross Formula in the hdPS Framework for Bias and Variance Estimation?

### 2024-09-13 Writeup/manuscript.Rmd
update2

title: Understanding the role of different proxy selection methods in High-Dimensional Propensity Score Framework - A Plasmode simulation study
titlerunning: High-Dimensional Propensity Score

@@

\textcolor{red}{*need to write what hdps is, and why alternatives are considered*}

\textcolor{blue}{\textbf{Introduction of hdPS}: The use of medico-administrative datasets in research is frequently criticized due to the potential for the lack of key confounders. Estimating treatment effects utilizing such data sources may be subject to residual confounding. To address this issue, the high-dimensional propensity score (hdPS) algorithm was introduced in 2009. The hdPS method systematically incorporates some information that are associated with the unmeasured or mismeasured confounders as proxies, with the goal of reducing residual confounding bias.}

\textcolor{blue}{\textbf{Significance of alternatives}: Since the introduction of hdPS algorithm, numerous extensions, rooted in machine learning and semi-parametric techniques, have been developed in order to improve the selection of proxy variables. The study aims to compare and evaluate the hdPS methods alongside the alternative proxy selection approaches to identify the ones that provide superior confounder adjustment, reduce bias, and enhance the precision of treatment effect estimation.}

**Aim**: The aim of this research is to systematically evaluate and compare different proxy selection methods within the context of high-dimensional propensity score (hdPS) analysis. Specifically, the study focuses on assessing how these methods, including alternative variable selection approaches, perform in selecting proxy variables for confounding adjustment compared to the traditional Bross method within the hdPS framework. We seek to determine whether these alternative methods offer superior performance in estimating treatment effects compared to the default Bross formula.

@@

**Motivating Example**: To explore the relationship between obesity and the risk of diabetes, we revisited this association using data from three cycles of the National Health and Nutrition Examination Survey (NHANES) covering the years 2013-2014, 2015-2016, and 2017-2018 [@karim2024high]. This analysis was informed by a thorough review of the existing literature [@saydah2014trends;@liu2013association;@kabadi2012joint;@ostchega2012abdominal]. To identify relevant covariates, we constructed a causal diagram based on established causal inference principles [@greenland1999causal]. The covariates included in our analysis were carefully selected and categorized into Demographic, Behavioral, Health History, Access-related, and Laboratory variables [@karim2024high]. While most of these variables were binary or categorical, the Laboratory variables were continuous.

**Plasmode simulation**: To rigorously assess the performance of the methods under consideration, we employed a plasmode simulation framework, which is particularly well-suited for reflecting real-world data structures and complexities [@franklin2014plasmode]. This approach was modeled after the analytic dataset derived from NHANES and involved resampling from the observed covariates and exposure information (i.e., obesity) without altering them. By mirroring key aspects of an actual epidemiological study, this simulation framework offers a significant advantage over traditional Monte Carlo simulations, which often rely on idealized assumptions.

@@

**True Data Generating Mechanism Used in Plasmode Simulation**: The primary goal of this study is to evaluate various variable selection methods under realistic conditions. To achieve this, we formulated the outcome data based on a specific model specification that incorporates both exposure and covariates, including investigator-specified and proxy variables. The model specification consists of three key components (See Appendices \S A and B for further details):

@@

3. *Inclusion of Proxy Variables*: Real-world studies often deal with unmeasured confounding, which researchers attempt to mitigate by adding proxy variables. However, when a large number of proxies are added, some may act as noise variables, contributing little to the analysis. To simulate this, we selected only those binary proxy covariates (referred to as recurrence covariates in hdPS terminology) that had a relative risk (RR) of less than 0.8 or greater than 1.2 concerning the outcome. Out of 143 \textcolor{brown}{\textbf{142}} proxy covariates, 94 met this criterion and were included in calculating a simple comorbidity burden measure. The remaining 49 \textcolor{brown}{\textbf{49? or 48?}} covariates were excluded from this calculation and considered noise. This comorbidity burden measure was then incorporated into our model specification for generating the plasmode data.

@@

\textcolor{red}{*check if the description of the methods look okay to you. Make sure the description matches with the analysis codes.*}

\textcolor{blue}{All looks good!}

The comparison between the data generation process and the analysis process reveals two key differences: (i) The data generation used transformed laboratory variables, whereas the analysis was conducted using only the original laboratory variables. (ii) The data generation employed a simple sum of selected proxy variables (sum of 94 proxy covariates), while the analysis included all proxy variables (143 \textcolor{brown}{\textbf{142}} binary proxies), with 49\textcolor{brown}{\textbf{49? or 48?}} of these acting as noise variables. These differences help us assess how the proxy variable selection methods handle model misspecification and the presence of noise variables.

>>>>

title: Is There a Competitive Advantage to Using Multivariate Statistical or Machine Learning Methods Over the Bross Formula in the hdPS Framework for Bias and Variance Estimation?
titlerunning: High-Dimensional Propensity Score

@@

**High-dimensional Propensity Score (hdPS) Algorithm**: In epidemiology, proxy variables are commonly used as substitutes for confounders that are difficult or impossible to measure directly, such as socioeconomic status, lifestyle factors, or health behaviors [@vanderweele2019principles]. The high-dimensional propensity score (hdPS) is a pharmacoepidemiological method designed to reduce confounding bias in large healthcare databases [@schneeweiss2009high]. Unlike traditional propensity score models that rely on investigator-specified or manually selected covariates, hdPS automatically ranks a wide array of proxy variables from healthcare records—such as diagnosis codes, medications, and procedures—using the Bross formula [@wyss2018erratum; @bross1966spurious]. The Bross formula ranks these variables based on their marginal associations with both exposure and outcome. These selected proxy variables serve as surrogates for unmeasured or poorly measured confounders, helping reduce bias in treatment effect estimates. The hdPS algorithm further refines the selection by prioritizing variables based on their prevalence and potential for confounding, defined by their association with both exposure and outcome [@schneeweiss2009high].

**Multivariate Machine Learning Extensions**: Although the Bross formula performs well in certain contexts, it has limitations in capturing complex interactions, nonlinearities, and higher-order associations between variables, especially in high-dimensional settings where it does not account for the multivariate structure of other covariates [@karim2024high; @karim2018can]. To address these limitations, multivariate machine learning methods such as LASSO, Elastic Net, and Random Forests have been applied within the hdPS framework. These methods are better suited for high-dimensional data, where they can more effectively handle complex relationships and improve the selection of proxy variables, thus enhancing the precision of treatment effect estimates [@karim2018can; @schneeweiss2017variable; @franklin2015regularized]. Simulation studies and empirical research have shown that these machine learning-based methods, or hybrid approaches combining the Bross formula with machine learning, can reduce confounding more effectively and increase efficiency compared to the Bross formula alone in certain settings [@karim2018can; @schneeweiss2017variable; @franklin2015regularized].

**Assessing the Simulation Performance**: Previous studies have often focused on Mean Squared Error (MSE) as a primary metric for evaluating the performance of hdPS and its machine learning extensions [@karim2018can; @franklin2015regularized]. However, in high-dimensional settings with singly robust methods, such as hdPS and machine learning approaches like LASSO, MSE may not always be the most reliable measure. MSE combines both bias and variance into a single metric, which makes interpretation challenging when variance estimation is unstable—a common issue in these methods. In contrast, coverage, which measures the proportion of confidence intervals that capture the true treatment effect, provides a more direct and meaningful assessment of a model's reliability. In realistic observational studies, where model misspecification is often inevitable, coverage—along with related metrics such as bias-eliminated coverage and relative error in standard error (which compares model-based SE with empirical SE)—can reveal whether confidence intervals or SEs are too narrow (underestimating uncertainty) or too wide (overestimating uncertainty). This insight is crucial in determining whether the model delivers valid estimates despite misspecification. Even a model with poor MSE but good coverage may still be valuable, as it produces realistic confidence intervals. By shifting the focus to coverage, rather than relying solely on MSE, we can achieve a more comprehensive understanding of method performance, especially in cases where unstable variance estimation might distort conclusions drawn from MSE alone.

**Aim**: This research aims to systematically evaluate and compare various proxy selection methods within the hdPS framework, using a diverse range of simulation performance metrics, including bias, MSE, and coverage. The study focuses on assessing how these alternative statistical and machine leartning methods perform in selecting proxy variables for confounding adjustment, compared to the traditional Bross formula.

@@

**Motivating Example**: We revisited the association between obesity and the risk of diabetes using data from three cycles of the National Health and Nutrition Examination Survey (NHANES) covering the years 2013-2014, 2015-2016, and 2017-2018 [@karim2024high]. To identify relevant investigator-specified covariates, we constructed a causal diagram based on literature [@saydah2014trends;@liu2013association;@kabadi2012joint;@ostchega2012abdominal] and established causal inference principles [@greenland1999causal]. The covariates included in our analysis were carefully selected and categorized into demographic, behavioral, health history, access-related, and laboratory variables. While most of these variables were binary or categorical, the Laboratory variables were continuous.

**Plasmode simulation**: To rigorously assess the performance of the methods under consideration, we employed a plasmode simulation framework, which is particularly well-suited for reflecting real-world data structures and complexities [@franklin2014plasmode]. This approach was modeled after the analytic dataset derived from NHANES and involved resampling from the observed covariates and exposure information (i.e., obesity) without altering them. By mirroring key aspects of an actual epidemiological study, this simulation framework offers a substantial advantage over traditional Monte Carlo simulations, which often rely on hypothetical assumptions.

@@

**True Data Generating Mechanism Used in Plasmode Simulation**: The primary goal of this plasmode simulation study is to evaluate various variable selection methods under realistic conditions. To achieve this, we formulated the outcome data based on a specific model specification that incorporates both exposure and covariates, including investigator-specified and proxy variables. The model specification consists of three key components (See Appendices \S A and B for further details):

@@

3. *Inclusion of Proxy Variables*: Real-world studies often deal with unmeasured confounding, which researchers attempt to mitigate by adding proxy variables. However, when a large number of proxies are added, some may act as noise variables, contributing little or nothing to the analysis. To simulate this, we selected only those binary proxy covariates (referred to as recurrence covariates in hdPS terminology) that had a relative risk (RR) of less than 0.8 or greater than 1.2 concerning the outcome. Out of 142 proxy covariates, 94 met this criterion and were included in calculating a simple comorbidity burden measure. The remaining 48 covariates were excluded from this calculation and considered noise. This comorbidity burden measure (one variable) was then incorporated into our model specification for generating the outcome data.

@@

The comparison between the data generation process and the analysis process reveals two key differences: (i) The data generation used transformed laboratory variables, whereas the analysis was conducted using only the original laboratory variables. (ii) The data generation employed a simple sum of selected proxy variables (sum of 94 proxy covariates), while the analysis included all proxy variables (142 binary proxies), with 48 of these acting as noise variables. These differences help us assess how the proxy variable selection methods handle model misspecification and the presence of noise variables.

### 2024-09-13 Writeup/manuscript.tex
update2

\title[High-Dimensional Propensity Score]{Understanding the role of
different proxy selection methods in High-Dimensional Propensity Score
Framework - A Plasmode simulation study}

@@

\textcolor{red}{*need to write what hdps is, and why alternatives are considered*}

\textcolor{blue}{\textbf{Introduction of hdPS}: The use of medico-administrative datasets in research is frequently criticized due to the potential for the lack of key confounders. Estimating treatment effects utilizing such data sources may be subject to residual confounding. To address this issue, the high-dimensional propensity score (hdPS) algorithm was introduced in 2009. The hdPS method systematically incorporates some information that are associated with the unmeasured or mismeasured confounders as proxies, with the goal of reducing residual confounding bias.}

\textcolor{blue}{\textbf{Significance of alternatives}: Since the introduction of hdPS algorithm, numerous extensions, rooted in machine learning and semi-parametric techniques, have been developed in order to improve the selection of proxy variables. The study aims to compare and evaluate the hdPS methods alongside the alternative proxy selection approaches to identify the ones that provide superior confounder adjustment, reduce bias, and enhance the precision of treatment effect estimation.}

\textbf{Aim}: The aim of this research is to systematically evaluate and
compare different proxy selection methods within the context of
high-dimensional propensity score (hdPS) analysis. Specifically, the
study focuses on assessing how these methods, including alternative
variable selection approaches, perform in selecting proxy variables for
confounding adjustment compared to the traditional Bross method within
the hdPS framework. We seek to determine whether these alternative
methods offer superior performance in estimating treatment effects
compared to the default Bross formula.

>>>>

\title[High-Dimensional Propensity Score]{Is There a Competitive
Advantage to Using Multivariate Statistical or Machine Learning Methods
Over the Bross Formula in the hdPS Framework for Bias and Variance
Estimation?}

@@

\textbf{High-dimensional Propensity Score (hdPS) Algorithm}: In
epidemiology, proxy variables are commonly used as substitutes for
confounders that are difficult or impossible to measure directly, such
as socioeconomic status, lifestyle factors, or health behaviors
\citep{vanderweele2019principles}. The high-dimensional propensity score
(hdPS) is a pharmacoepidemiological method designed to reduce
confounding bias in large healthcare databases
\citep{schneeweiss2009high}. Unlike traditional propensity score models
that rely on investigator-specified or manually selected covariates,
hdPS automatically ranks a wide array of proxy variables from healthcare
records---such as diagnosis codes, medications, and procedures---using
the Bross formula \citep{wyss2018erratum, bross1966spurious}. The Bross
formula ranks these variables based on their marginal associations with
both exposure and outcome. These selected proxy variables serve as
surrogates for unmeasured or poorly measured confounders, helping reduce
bias in treatment effect estimates. The hdPS algorithm further refines
the selection by prioritizing variables based on their prevalence and
potential for confounding, defined by their association with both
exposure and outcome \citep{schneeweiss2009high}.

\textbf{Multivariate Machine Learning Extensions}: Although the Bross
formula performs well in certain contexts, it has limitations in
capturing complex interactions, nonlinearities, and higher-order
associations between variables, especially in high-dimensional settings
where it does not account for the multivariate structure of other
covariates \citep{karim2024high, karim2018can}. To address these
limitations, multivariate machine learning methods such as LASSO,
Elastic Net, and Random Forests have been applied within the hdPS
framework. These methods are better suited for high-dimensional data,
where they can more effectively handle complex relationships and improve
the selection of proxy variables, thus enhancing the precision of
treatment effect estimates
\citep{karim2018can, schneeweiss2017variable, franklin2015regularized}.
Simulation studies and empirical research have shown that these machine
learning-based methods, or hybrid approaches combining the Bross formula
with machine learning, can reduce confounding more effectively and
increase efficiency compared to the Bross formula alone in certain
settings
\citep{karim2018can, schneeweiss2017variable, franklin2015regularized}.

\textbf{Assessing the Simulation Performance}: Previous studies have
often focused on Mean Squared Error (MSE) as a primary metric for
evaluating the performance of hdPS and its machine learning extensions
\citep{karim2018can, franklin2015regularized}. However, in
high-dimensional settings with singly robust methods, such as hdPS and
machine learning approaches like LASSO, MSE may not always be the most
reliable measure. MSE combines both bias and variance into a single
metric, which makes interpretation challenging when variance estimation
is unstable---a common issue in these methods. In contrast, coverage,
which measures the proportion of confidence intervals that capture the
true treatment effect, provides a more direct and meaningful assessment
of a model's reliability. In realistic observational studies, where
model misspecification is often inevitable, coverage---along with
related metrics such as bias-eliminated coverage and relative error in
standard error (which compares model-based SE with empirical SE)---can
reveal whether confidence intervals or SEs are too narrow
(underestimating uncertainty) or too wide (overestimating uncertainty).
This insight is crucial in determining whether the model delivers valid
estimates despite misspecification. Even a model with poor MSE but good
coverage may still be valuable, as it produces realistic confidence
intervals. By shifting the focus to coverage, rather than relying solely
on MSE, we can achieve a more comprehensive understanding of method
performance, especially in cases where unstable variance estimation
might distort conclusions drawn from MSE alone.

\textbf{Aim}: This research aims to systematically evaluate and compare
various proxy selection methods within the hdPS framework, using a
diverse range of simulation performance metrics, including bias, MSE,
and coverage. The study focuses on assessing how these alternative
statistical and machine leartning methods perform in selecting proxy
variables for confounding adjustment, compared to the traditional Bross
formula.
